---
title: "Pre-Analysis Plan Version 1"
author: "HEGSRR"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  markdown:
    wrap: sentence
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../docs") })
nocite: '@*'
bibliography: "../../software.bib"
---

# Abstract

This study is a *reproduction* of: Tuholske, C., Lynch, V.D., Spriggs, R.
et al.
Hazardous heat exposure among incarcerated people in the United States.
Nat Sustain 7, 394–398 (2024).
<https://doi.org/10.1038/s41893-024-01293-y> [Git Repository](https://github.com/sparklabnyc/temperature_prisons_united_states_2024)

By reproducing the exploratory data analysis done in Tuholske et al. (2024), we seek to...
*1.* Identify the temporal resolution at which authors use population data to calculate population weighted hazardous heat days *2.* Understand the population weighting mechanism in state-level hazardous heat calculations.
*3.* Evaluate the effectiveness of the author's methods compared to similar research.

# Study metadata

-   `Key words`: Comma-separated list of keywords (tags) for searchability. Geographers often use one or two keywords each for: theory, geographic context, and methods.
-   `Subject`: Social and Behavioral Sciences: Geography: Geographic Information Sciences, Human Geography, Nature and Society Relations
-   `Date created`: 04/14/25
-   `Date modified`: 04/21/25
-   `Spatial Coverage`: United States Lower 48
-   `Spatial Resolution`: Carceral facility points and States
-   `Spatial Reference System`: Specify the geographic or projected coordinate system for the study, e.g. EPSG:4326
-   `Temporal Coverage`: 1982-2020
-   `Temporal Resolution`: 1 year

## Original study spatio-temporal metadata

-   `Spatial Coverage`: United States Lower 48
-   `Spatial Resolution`: Carceral facility points and States
-   `Spatial Reference System`: spatial reference system of original study
-   `Temporal Coverage`: 1982-2020
-   `Temporal Resolution`: 1 year

# Study design

This study is a **reproduction study**, and the original study is more **exploratory** in design.
The primary objective for exploration through research and analysis investigated in the original study was exposure to hazardous heat in carceral facilities in the continental U.S.
The authors wanted to examine how exposure to hazardous heat changed over time from 1982-2020, as well as how exposure within carceral facilities compared to exposure in the rest of the state.
In general, determining the spatial distribution of carceral facilities with higher levels of hazardous heat exposure was also an objective of the original paper

# Materials and procedure

## Computational environment

```{r environment-setup, include = FALSE}
# record all the packages you are using here
# this includes any calls to library(), require(),
# and double colons such as here::i_am()
packages <- c("tidyverse", "here", "markdown", "htmltools")

# force all conflicts to become errors
# if you load dplyr and use filter(), R has to guess whether you mean dplyr::filter() or stats::filter()
# the conflicted package forces you to be explicit about this
# disable at your own peril
# https://conflicted.r-lib.org/
require(conflicted)

# load and install required packages
# https://groundhogr.com/
if (!require(groundhog)) {
  install.packages("groundhog")
  require(groundhog)
}

# this date will be used to determine the versions of R and your packages
# it is best practice to keep R and its packages up to date
groundhog.day <- "2025-02-01"

# this replaces any library() or require() calls
groundhog.library(packages, groundhog.day,
                  tolerate.R.version='4.4.2')
# you may need to install a correct version of R
# you may need to respond OK in the console to permit groundhog to install packages
# you may need to restart R and rerun this code to load installed packages
# In RStudio, restart r with Session -> Restart Session

# record the R processing environment
# alternatively, use devtools::session_info() for better results
writeLines(
  capture.output(sessionInfo()),
  here("procedure", "environment", paste0("r-environment-", Sys.Date(), ".txt"))
)

# save package citations
knitr::write_bib(c(packages, "base"), file = here("software.bib"))

# set up default knitr parameters
# https://yihui.org/knitr/options/
knitr::opts_chunk$set(
  echo = FALSE, # Show outputs, but not code. Change to TRUE to show code as well
  fig.retina = 4,
  fig.width = 8,
  fig.path = paste0(here("results", "figures"), "/")
)
```

## Data and variables

### Population Data

```{r}
includeMarkdown(
  here("data", "metadata", "metadata_population.md")
)
```

### Prison Boundaries

```{r}
includeMarkdown(
  here("data", "metadata", "metadata_Prison_Boundaries.md")
)
```

### State Boundaries

```{r}
includeMarkdown(
  here("data", "metadata", "metadata_statesshp.md")
)
```

### WBGT Data - Prison Level

```{r}
includeMarkdown(
  here("data", "metadata", "metadata_wbgt_prison.md")
)
```

### WBGT Data - State Level

```{r}
includeMarkdown(
  here("data", "metadata", "metadata_wbgt_state.md")
)
```

## Prior observations

At the time of this pre analysis plan, we have the derived data to work off of, we have neither visualized or analyzed prison data or WBGTmax temperature data before.

This study 

## Bias and threats to validity

Given the research design and primary data to be collected and/or secondary data to be used, discuss common threats to validity and the approach to mitigating those threats, with an emphasis on geographic threats to validity.

These include: - uneven primary data collection due to geographic inaccessibility or other constraints - multiple hypothesis testing - edge or boundary effects - the modifiable areal unit problem - nonstationarity - spatial dependence or autocorrelation - temporal dependence or autocorrelation - spatial scale dependency - spatial anisotropies - confusion of spatial and a-spatial causation - ecological fallacy - uncertainty e.g. from spatial disaggregation, anonymization, differential privacy

## Data transformations

Planned deviation: we will not attempt x because the data is .... therefore we will skip to ...

Describe data transformations conducted by the authors that are not feasible for reproduction. 

Explain what we believe the authors did for WBGTmax grid - document as data source we do not have access to, and its lineage

Give readers a sense of the workflow that we are not doing/unable to do

#### Step 1: Join author-provided WBGTmax by day grid data to author-provided carceral facility point data

Use `st_join()`

*Result:* Rds table of WBGTmax by day by prison

#### Step 2: Filter result by days when WBGTmax exceeded 28 degrees C

#### Step 3: Group by carceral facility type and year

Count to produce summary of days exceeded per year by facility

*Result:* Rds table with variables\
- prison facility - facility type - prison population - n days exceeding 28 degrees - year

#### Step 4: Begin with result from Step 3 from reproduction of Fig. 1

#### Step 5: Spatial join county population by year data

*Result:* Rds table with variables\
- prison facility - facility type - prison population - county - population - n days exceeding 28 degrees - year

#### Step 6: Population-weighted aggregation

Aggregate data into states

weighted sum of days exceeded across all counties of the state

sum of days exceeded multiplied by (Ratio of county population / state population)

## Planned deviation for reproduction: Investigating temperature threshold

Repeat workflow for reproducing figures 1 and 2, instead filtering for days when WBGTmax exceeded 29.4 degrees C (85 degrees F standard informed by other literature)

## Planned deviation for reproduction: Investigating sources of uncertainy/error/bias

### How many open facilities had a population of -999? (Potential source of uncertainy)

#### Step 7: Select facilities with population of -999 from author-provided carceral facilities data.

Report number of facilities compared to authors' number.

## Analysis

If anything - Fig 2c - comparison of prisons to state

# Results

## Create Fig. 1a and b using results from step 3

## Create Fig. 2a, b and c using results from step 6

## Map result from step 7 to examine planned deviation

Move steps that create maps and graphs into results

Present figures (reproduced from original study and planned deviations).

# Discussion

what are the implications of us being able to recreate or not recreate the figures? why does it matter for this to be reproducible

mention groundhog usage for sustainable reproduction

Describe how the results are to be interpreted *vis a vis* each hypothesis or research question.
Research that maximum daily doesn't matter as much for heat stress - long stretches of nighttime lows are more serious

# Integrity Statement

This is the first version of our pre-analysis plan. Any deviations in our workflow will be documented as unplanned deviations.

# Acknowledgements

This report is based upon the template for Reproducible and Replicable Research in Human-Environment and Geographical Sciences, [DOI:[10.17605/OSF.IO/W29MQ](DOI:%5B10.17605/OSF.IO/W29MQ){.uri}](https://doi.org/10.17605/OSF.IO/W29MQ)

## References

> Kedron, P., & Holler, J.
> (2023).
> Template for Reproducible and Replicable Research in Human-Environment and Geographical Sciences.
> <https://doi.org/10.17605/OSF.IO/W29MQ>
